---
title: 'Prediction of 2020 US Election: Trumping winning 63%'
author: "Yuxuan Liu, Yuxuan Lin, Yangle Shang, Ziqin Zhou"
date: '2020-11-02'
output:
  html_document:
    df_print: paged
  pdf_document: default
---
```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(pROC)
library(arm)
library(reshape2)
library(ggpubr)
# Loading in the cleaned survey Data
reduced_survey <- read.csv("survey_data1.csv")
# Loading in the cleaned census Data
reduced_census <- read.csv("census_data1.csv")
```

Code and data supporting this analysis is available at:
https://github.com/shang0908/Prediction-of-2020-US-election

# Model
```{r,echo=F, include=FALSE, message=F}
# Creating the Model
model_logit_1 <- glmer(vote_trump ~ (1 + education|state) + race + age + labforce + sex,
                       data = reduced_survey, 
                       family = binomial)
model_logit_2 <- glmer(vote_biden ~ (1 + education|state) + race + age + labforce + sex,
                       data = reduced_survey,
                       family = binomial)
```

## Model Specifics
Here, we decide to build random coefficient logistic regression models in R about Trump and Biden's vote count by frequentest approach. By looking at the past presidential elections, we noticed that there are some states that voted for neither of the two most popular candidates. Thus the prediction on vote count will be more precise with two separate model on Biden and Trump than with one. Also, we choose to change the numeric variable age into age group because we want to know the preference of the young, mid-aged and elderly. The two models for Trump and Biden's vote count are shown below.
$$P(Y_{i}={Trump}) =logit^{-1}(\beta_{0} + X^{edu}_{state[ij]} +\beta^{race}_{ij}+\beta^{age}_{ij}+\beta^{labforce}_{ij} + \beta^{sex}_{ij} )$$
$$P(Y_{i}={Biden}) =logit^{-1}(\beta_{0} + X^{edu}_{state[ij]} +\beta^{race}_{ij}+\beta^{age}_{ij}+\beta^{labforce}_{ij} + \beta^{sex}_{ij})$$
where $\beta_{0}$ is the fixed base intercept and $\beta^{race}_{ij}$, $\beta^{agr}_{ij}$, $\beta^{labforce}_{ij}$, $\beta^{sex}_{ij}$ represent coefficients associated with each categorical variable. The $X^{edu}_{state[ij]}$ shows the random effect for all states for each education level. The coeffcient matrix of random effect gives the slope for all the states for each education level. The subscript $_{ij}$ indicates the $j$ state to which the education$i$ respondent belongs. Except for the variable race, p-values of all of the other predictors are significant that are less than 0.05, indicating we could reject the null hypothesis.

```{r,echo=F,message=F}
#check the model
prob1 <- predict(model_logit_1, type=c('response')) #trump predict
prob2 <- predict(model_logit_2, type=c('response')) #biden predict

roc1 <- roc(reduced_survey$vote_2020, prob1)
auc(roc1)
plot(roc1, auc.polygon=TRUE, print.auc = TRUE, asp = NA) #draw the roc for Trump

roc2 <- roc(reduced_survey$vote_2020, prob2)
auc(roc2)
```

The receiving operating characteristic (ROC) curve will be use to check the diagnostic ability of our models. We will obtain the ROC curve by plotting the sensitivity (true positive rage) against 1-specificity (true negative rate). The AUC (area under ROC curve) values is respectively 0.6865 for "Trump model" and 0.6065 for "Biden model" which means that there is 68.65%/60.65% chance to discriminate between Trump/Biden gets the vote or not. Then, we use the binned plot to check the residuals. Since the gray lines represent ±2SE bands, which we would expect to contain about 85% of survey data about vote for Trump and Biden. Since the majority of the fitted values fall within the SE bands for both models, the models is reasonable.

```{r,echo=F,message=F}

binnedplot(fitted(model_logit_1), residuals(model_logit_1, type = "response"),
           nclass = NULL, xlab = "Expected Values", ylab = "Average residual", 
           main = "Binned residual plot for Trump", 
           cex.pts = 0.8, 
           col.pts = 1, 
           col.int = "gray") #draw the binnedplot for Trump model

```

## Post-Stratification 
Post-stratification is a method used to adjust the sampling weights. It could increase the representativeness of the sample, so we have greater confidence of our result. In order to correct for sampling bias, we need to perform a post-stratification analysis. Since "Winner take all" and each state is represented different number of electors, we create cells based off different states. Using the model described in the previous sub-section, we estimate the proportion of voters in each state. Then we weight each proportion estimate in each state by the population size of that bin respectively. Finally, we compare the proportion of voting for Biden and Trump in each state. Due to "winner-take-all", the winner is the one with higher proportion and he will take all electoral votes of that states. By summing them up for each person, we get the result.
```{r,echo=F, message=F}
reduced_census$trumpestimate <-
  model_logit_1 %>%
  predict(newdata = reduced_census, type = "response") #estimate chance of voting for trump

reduced_census$bidenestimate <-
  model_logit_2 %>%
  predict(newdata = reduced_census, type = "response") #estimate chance of voting for biden

vote_2020_pred <- ifelse(reduced_census$bidenestimate > reduced_census$trumpestimate, "Joe Biden", "Donald Trump") #the final vote

predictresult <- cbind(reduced_census, vote_2020_pred) #combine the result 

predictresult$trump_votes <- ifelse(predictresult$vote_2020_pred == "Donald Trump", predictresult$perwt, 0) #add the perwt to get number of people that the respondents represent

predictresult$biden_votes <- ifelse(predictresult$vote_2020_pred == "Joe Biden", predictresult$perwt, 0) #add the perwt to get number of people that the respondents represent 
 
predictresult %>%
  group_by(state) %>%
  summarise(Trump = sum(trump_votes), Biden = sum(biden_votes)) -> predicted_states #group by state and find the votes they get

```

# Results
In the 538 polls of electoral colleges, Trump will win 339 votes. As a result, our estimate of the proportion of voters for Donald Trump is 0.63, based off our post-stratification analysis on the random coefficient logistic regression model above. In particular, Democrats states include Massachusetts, Maryland, Michigan and Oregon. On the contrary, Alabama, Mississippi, Kansas and Idaho are loyal to the Republican party. Our predicted results tell us all others except for Oregon vote for their original party they support with a enormous gap. But the voting is not too dramatic in Oregon, Republicans win with just 7% of the vote.
```{r,echo=F,message=F}
predicted_states$winner<-ifelse(predicted_states$Trump > predicted_states$Biden,
                                "Donald Trump","Joe Biden") #more get all
predicted_states <- predicted_states %>% 
  mutate(electoral_votes =
           case_when(state=="CA"~55,state=="TX"~38,state=="FL"~29,state=="NY"~29,state=="IL"~20,
                     state=="PA"~20,state=="OH"~18,state=="GA"~16,state=="MI"~16,state=="NC"~15,
                     state=="NJ"~14,state=="VA"~13,state=="WA"~12,state=="AZ"~11,state=="IN"~11,
                     state=="MA"~11,state=="TN"~11,state=="MD"~10,state=="MN"~10,state=="MO"~10,
                     state=="WI"~10,state=="AL"~9,state=="CO"~9,state=="SC"~9,state=="KY"~8,
                     state=="LA"~8,state=="CT"~7,state=="OK"~7,state=="OR"~7,state=="AR"~6,
                     state=="IA"~6,state=="KS"~6,state=="MS"~6,state=="NV"~6,state=="UT"~6,
                     state=="NE"~5,state=="NM"~5,state=="WV"~5,state=="HI"~4,state=="ID"~4,
                     state=="ME"~4,state=="NH"~4,state=="RI"~4,state=="AK"~3,state=="DE"~3,
                     state=="MT"~3,state=="ND"~3,state=="SD"~3,state=="VT"~3,state=="WY"~3,
                     state=="DC"~3)) 
#get the elevtoral votes of each state by google(reference)

           
predicted_states %>% 
  group_by(winner) %>% 
  summarise(total_votes = sum(electoral_votes)) -> election_result #sum the electorical votes they get

election_result #show the result
```

# Discussion
The survey data and the census data are collected from the Democracy Fund + UCLA Nationscape ‘Full Data Set’ and IPUMS USA website respectively. In order to predict 2020 US election and to observe which variables obtain significant relation with the response variable(vote_2020), we construct the intention variable and perform a clean process for changing in demographic variables. However, the survey data could not represent the whole population(as shown in the weakness). Therefore, we use the MRPs to correct the bias. We build the Multilevel logistic regression model with random effect on the individual level survey data. Then, we divide the poll into 51 cells by states. In the post-stratification process, we count the actual effective votes by electoral college in each state. The candidate who gets the most votes gets the votes of all the state's electors.
The result shows Trump will win 2020 US general election, the Democrats party being 0.63 among 538 polls of electoral colleges. A fall-off in Black turnout was one of the key reasons Clinton lost in 2016. According to data site FiveThirtyEight, African Americans remain firmly in the Democratic camp. That's may be partly explain that Trump are likely to win again. Investors think the election outcome may influence the stock market.It will lead to portfolio winners and losers for that election uncertainty.

## Weaknesses & next step
```{r, echo=F,message=F}
totalsurvey = nrow(reduced_survey)
totalcensus = nrow(reduced_census)  #calculate the total of census and survey
reduced_survey %>%
  group_by(labforce) %>%
  summarise(count = n(), surveyratio = n()/totalsurvey) -> labforce #count the labforce of survey
reduced_census %>%
  group_by(labforce) %>%
  summarise(count = n(), ratiolabforce = n()/totalcensus) -> censuslabforce #count the labforce of census
cbind(labforce, censusratio = censuslabforce$ratiolabforce) -> labforce
labforce = as.matrix(labforce[, 3:4])
rownames(labforce) = c("no, not in the labor force", "yes, in the labor force")
colnames(labforce) = c("survey", "census") #create the matrix
labforceplot <-ggplot(melt(labforce), aes(x = Var1, y = value, col = Var2), ylab = "ratio") +
  ggtitle("the comparison of labor force ratio in survey and census") +
  theme(plot.title = element_text(hjust = 0.5)) + 
  geom_point() +
  geom_line(aes(group = Var2), size = 1) +
  ylim(0, 1) +
  coord_fixed(ratio = 0.8) #draw the plot
```
Firstly, despite the large sample size, the pool of survey respondents is far from being representative of the voting population. we compares the demographic composition of survey to that of the census. We plot the two most striking differences variables: labforce and education below.

```{r, echo=F,message=F}
reduced_survey %>%
  group_by(education) %>%
  summarise(count = n(), surveyratio = n()/totalsurvey) -> education
reduced_census %>%
  group_by(education) %>%
  summarise(count = n(), ratioeducation = n()/totalcensus) -> censuseducation
cbind(education, censusratio = censuseducation$ratioeducation) -> education
education = as.matrix(education[, 3:4])
rownames(education) = c("college graduation and above", "Don't graduate from HS", 
                "High school graduate", "some college")
colnames(education) = c("survey", "census")
educationplot <- ggplot(melt(education), aes(x = Var1, y = value, col = Var2), ylab = "ratio") +
  ggtitle("the comparison of education ratio in survey and census") +
  theme(plot.title = element_text(hjust = 0.5)) + 
  geom_point() +
  geom_line(aes(group = Var2), size = 1) +
  ylim(0, 0.5) +
  coord_fixed(ratio = 3) +
  theme(axis.text.x = element_text(vjust = 1,
                                   hjust = 1,
                                   size = 7))
ggarrange(educationplot, labforceplot,
          ncol = 1, nrow = 2)
```

We could observe that people who have taken college and above account for the largest proportion of about 40% in the survey, compared to only 30% college and above graduate in the census. High school graduate make up 28% of the survey but only 18% of the census. 
Moreover, people in the labor force dominated the survey population: they comprises about 75% of the survey dataset. By contrast, the proportion of "in the labor force" and "not in the labor force" is much more close to each other. We would like to make our survey data more representative since it is the base of forecasting.
Secondly, we just partition our data into 51 cells to simplify and reduce the running time at the cost of accuracy. It does not correct the model perfectly. We could partition our data and involve more cells in the future stage.
What's more, according to Zalesny(1985), economic factors are very important in vote preference. However, our dataset does not include any variable related to economic factors since the census dataset just has personal income while survey has household income. We will include it in the next stage.
Lastly, we will compare our estimation with the actual election results and do a post-hoc analysis of how to better improve estimation in future elections.

# References
Data resources:
Nationscape Methodology and Representativeness Assessment. Democracy Fund + UCLA Nationscape, 30 Oct. 2020, www.voterstudygroup.org/publication/nationscape-data-set. 

“2018 5-year ACS”. IPUMS USA, American Community Survey and Puerto Rican Community Survey, 2020,
https://usa.ipums.org/usa/acs_multyr.shtml

“Ignore Files from Version Control.” Ignore Files from Version Control – Git Using RStudio,
https://carpentries-incubator.github.io/git-Rstudio-course/02-ignore/index.html

Langsæther, Peter  Egge. “Why Does Class Affect Voting?” Democratic Audit, 2 July 2018, www.democraticaudit.com/2018/07/02/why-does-class-affect-voting/. 

Luthra, Shefali. “These 'pro-Life' Women Won't Vote for Trump. Here's Why.” USA Today, Gannett Satellite Information Network, 30 Oct. 2020, www.usatoday.com/story/news/politics/elections/2020/10/30/some-abortion-foes-redefine-pro-life-no-longer-support-trump/3755235001/. 

Sarkar, Kanishka. “US Election 2020- Donald Trump vs Joe Biden: A Look at Republican and Democratic Parties' Vision for America.” Hindustan Times, 30 Aug. 2020, www.hindustantimes.com/world-news/us-election-2020-donald-trump-vs-joe-biden-a-look-at-republican-and-democratic-party-s-vision-for-america/story-fcIoqYn3KM2H8wgjfvjSFP.html. 

Prokop, Andrew. “Why the Electoral College Is the Absolute Worst, &nbsp;Explained.” Vox, Voxmedia, 7 Nov. 2016, www.vox.com/policy-and-politics/2016/11/7/12315574/electoral-college-explained-presidential-elections-2016. 

Lauren Leatherby, Journalist's Resource November 6. “The Role of Race in Voter Turnout.” Journalist's Resource, 16 Nov. 2016,
https://journalistsresource.org/studies/politics/elections/minority-voter-turnout-research/.

“List of Political Parties in the United States.” Ballotpedia, https://ballotpedia.org/List_of_political_parties_in_the_United_States 

Zalesny, M. D. (1985). Comparison of economic and noneconomic factors in predicting faculty vote preference in a union representation election. Journal of Applied Psychology, 70(2), 243–256. https://doi.org/10.1037/0021-9010.70.2.243

Trump Is Losing Ground With White Voters But Gaining Among Black And Hispanic Americans, https://fivethirtyeight.com/features/trump-is-losing-ground-with-white-voters-but-gaining-among-black-and-hispanic-americans/

pROC package
https://www.rdocumentation.org/packages/pROC/versions/1.16.2

Understanding ROC-AUC curve
https://towardsdatascience.com/understanding-auc-roc-curve-68b2303cc9c5

arm Package
https://www.rdocumentation.org/packages/arm/versions/1.11-2

reshape2 Package
https://www.rdocumentation.org/packages/reshape2/versions/1.4.4

ggpubr Package
https://www.rdocumentation.org/packages/ggpubr/versions/0.1.1

